---
title: "PH125.9x Capstone Abalone Project"
author: "Michael W. Jones"
date: "8/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
# if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("gridExtra", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")
if(!require(tinytex)) tinytex::install_tinytex()

if(!require(PivotalR)){install.packages("PivotalR")}
if(!require(Metrics)){install.packages("Metrics")}
if(!require(gbm)){install.packages("gbm")}
if(!require(gam)){install.packages("gam")}
if(!require(splines)){install.packages("splines")}
if(!require(foreach)){install.packages("foreach")}
if(!require(gridExtra)){install.packages("gridExtra")}
if(!require(corrplot)){install.packages("corrplot")}
if(!require(doMC)){
  install.packages("doMC",
                   repos="http://R-Forge.R-project.org")}

doMC::registerDoMC(cores=4)

```


# Introduction
<!-- (an introduction/overview/executive summary section that describes the dataset and variables, and summarizes the goal of the project and key steps that were performed) -->

## Executive Summary
Abalone, any of several marine snails, constituting the genus Haliotis and family Haliotidae in the subclass Prosobranchia (class Gastropoda), in which the shell has a row of holes on its outer surface. Abalones are found in warm seas worldwide.

The dishlike shell is perforated near one edge by a single row of small holes that become progressively filled during the animal’s growth; the last five to nine holes remain open to serve as outlets for the snail’s waste products. 

The shell’s lustrous, iridescent interior is used in the manufacture of ornaments. The large muscular foot of the abalone is eaten as a delicacy in several countries. 

Depending on the species, abalones usually range from 10 to 25 cm (4 to 10 inches) across and up to 7.5 cm in depth. About 50 species have been described. The largest abalone is the 30-cm red abalone (H. rufescens) of the western coast of the United States. H. rufescens and several other species are raised commercially in abalone farms, particularly in Australia, China, Japan, and along the western coast of the United States. Commercial fisheries for abalones exist in California, Mexico, Japan, and South Africa. (Britannica, 2020)

The age of abalone is determined by cutting the shell through the cone, staining it, and counting the number of rings through a microscope - a tedious and time-consuming task. Other measurements, which are easier to obtain, may be used to predict the age. The age is calculated as 1.5 plus the number of rings.

Additional information, such as weather patterns and location (hence food availability) may be required to accurately predict age (i.e. rings).

## Overview
This project uses the information from the abalone dataset in the PivotalR library to identify the best model for predicting rings. 

```{r table description, echo=FALSE}

abalone_vars <- 
  read.csv("abalone_vars.csv")

names(abalone_vars) <- c("Name", "Data.Type",
                         "Measure", "Description")

abalone_vars %>% knitr::kable() 
rm(abalone_vars)

```

## Key Steps
The abalone dataset is loaded from the PivotalR library.  It is then divided into training and testing data sets, 90% and 10% respectively.  

In order to determine the best model for predicting rings, and hence the highest accuracy, both regression and classification analyses were done.  Models used for regression were:  lm, glm, knn, svmLinear, rpart, gamLoess, treebag, gbm, and rf.  The classification models were:  knn, lda, qda, naive_bayes, svmLinear, svmRadial, gamLoess, multinom, and rf.  Classification was done by grouping rings into age groups:  juvenile, adult, and senior.  The best models (least RMSE / highest accuracy) were chosen from each set of analysis.  The results and conclusions from the analysis are reported.

## Data Prep
The abalone dataset is obtained from the PivotalR library.  Sex was converted to a number: I = 0; F = 1; and M = 2 for both regression and classification analysis.  For classification, rings were grouped as juvenile (1-7); adult (8-11); and senior (12-29).  The grouping was based on a summary() of rings using the 1st quartile for juvenile, between the 1st and 3rd quartile for adult, and senior the remainder.

```{r summary(rings)}
summary(abalone$rings)
```

```{r data prep, include=F}

##################################################
##
##   Data Prep: load PivotalR::abalone dataset
##
##################################################

if(exists("abalone")){rm(list = ls())}
data(abalone)

# remove id column since not needed
abalone <- abalone %>% select(-id)

##
## data prep for regression
##

#create numeric sex column 
abalone_r <- abalone %>%
  mutate(sex_num =
           case_when(
             sex == "I" ~ 0,
             sex == "F" ~ 1,
             sex == "M" ~ 2
           )) 

# drop sex 
abalone_r <- abalone_r %>% select(-sex)

# reorder so sex_num is at top where sex was
abalone_r <- cbind(abalone_r[9], abalone_r[1:8])

# # # factor rings
# abalone_r$rings <- factor(abalone_r$rings)

save(abalone_r, file = "abalone_r.Rdata")

##
##  split_data() into test and train - 90% train
##

set.seed(2020)
idx <-
  createDataPartition(
    y = abalone$rings,
    times = 1,
    p = 0.1,
    list = FALSE
  )

train_set_r <- abalone_r[-idx,]
test_set_r <- abalone_r[idx,]



##
## data prep for classification
##

# based on rings breakdown, histogram and summary,
# classify age groups using ring quartiles:
# junior, adult, senior

abalone_c <- abalone_r %>%
  mutate(rgrp = case_when(
    between(rings, 1, 7) ~ "juvenile",
    between(rings, 8, 11) ~ "adult",
    between(rings, 12, 29) ~ "senior"
  ))

# remove rings 
abalone_c <- abalone_c %>% select(-rings)

```

# Analysis Methods
<!-- (a methods/analysis section that explains the process and techniques used, including data cleaning, data exploration and visualization, insights gained, and your modeling approaches (you must use at least two different models or algorithms);) -->

Data analysis begins by visualizing the data. The abalone data is visualized in several different ways, including: looking at a sampling of the data, the data structure, summary, correlation, density plots, histograms, box plots, pairs plot, and correlation plot.

Subsequently, regression analysis is performed using nine different model methods (lm, glm, knn, svmLinear, rpart, gamLoess, treebag, gbm, and rf).  After grouping the rings as juvenile, adult, and senior, classification analysis was done using nine model methods as well (knn, lda, qda, naive_bayes, svmLinear, svmRadial, gamLoess, multinom, and rf).  

The results of the analysis are reported for each analysis type - regression and classification, identifying the best model for each.

## Visualizing the data

### Data Sample
```{r data sample}
head(abalone, 3)
```

### Structure
```{r str(abalone)}
str(abalone)
```

### Summary
```{r summary(abalone_r)}
summary(abalone_r)
```

### Correlation data
```{r cor(abalone_r)}
cor(abalone_r[9], abalone_r[-9])
```


<!-- function for plotting -->
```{r plot_data function, include=F}
# function to create list of plots by geom
plot_data <- 
  function(x_data, x_col, x_geom, x_parm = "") {
  fx <- "qplot(x_data[,x_col], 
               geom = x_geom, 
               xlab = x_col)"
  fx_parm <- paste0(fx, x_parm)
  eval(parse(text = fx_parm))
}
```

\pagebreak
### Density plots
```{r density plots, echo=F}
x_geom <- "density"
plots <- lapply(colnames(abalone[-1]),
                plot_data, 
                x_data = abalone[-1], 
                x_geom = x_geom, 
                x_parm = "")
grid.arrange(grobs = plots, ncol = 2, top = x_geom)
```

\pagebreak
### Histograms
```{r histograms, echo=F}
# histograms each attribute
x_geom <- "histogram"
plots <- lapply(colnames(abalone[-1]),
                plot_data, 
                x_data = abalone[-1], 
                x_geom = x_geom, 
                x_parm = "")
suppressMessages(message(
  grid.arrange(grobs = plots, 
               ncol = 2, 
               top = x_geom)))
```

\pagebreak
### Box plots
```{r boxplots, echo=F}
# boxplots
x_geom <- "boxplot"
plots <- lapply(colnames(abalone[-1]),
                plot_data, 
                x_data = abalone[-1], 
                x_geom = x_geom, 
                x_parm = "")
grid.arrange(grobs = plots, ncol = 2, top = x_geom)
rm(x_geom, plots)
```

### Pairs plot
```{r pairs plots, echo=F}
# pairs matrix scatterplot
pairs(abalone_r)

```

### Correlation graphic
```{r Correlation graphic, echo=F}
# correlation plot color heat map
correlations <- cor(abalone[-1])
corrplot(correlations, 
       method="circle") 

```

<!-- ```{r cleanup 1, include=F} -->
<!-- ## abalone dataset since no longer used -->
<!-- rm(abalone) -->
<!-- ``` -->

\pagebreak
## Analysis
Regression and classification analyses are done for multiple models using the following basic code:

```{r dont_show, eval=F}

fits <- lapply(models, function(model) {
  print(model)
  set.seed(seed)
  train(
    rings ~ .,
    data = train_set_r,
    method = model,
    metric = metric,
    trControl = control
  )
})

```



There is some variation in the code depending upon whether regression or classification analysis.



## Regression Analysis
```{r regression analysis, include=F}
## Apply multiple models
models <- c("lm",
            "glm",
            "knn",
            "svmLinear",
            "rpart",
            "gamLoess",
            "treebag",
            "gbm",
            "rf")

# initialize variables
control <- trainControl(
  method="repeatedcv",
  number=10,
  repeats=3)
metric <- "RMSE"
seed <- 3

# run regression models if not exists
if(!file.exists("regress_fits.Rdata")){
regress_fits <- lapply(models, function(model) {
  print(model)
  set.seed(seed)
  train(
    rings ~ .,
    data = train_set_r,
    method = model,
    metric = metric,
    trControl = control
  )
})
names(regress_fits) <- models
save(regress_fits, file = "regress_fits.Rdata")
}

if(!exists("regress_fits")){
  load(file = "regress_fits.Rdata")}

results <- map_df(regress_fits, function(model) {
  method <- model$method
  RMSE <- model$results$RMSE %>% min()
  return(list("method" = method, "RMSE" = RMSE))
})

print_regress_results <- results %>%
  as_tibble() %>%
  arrange(RMSE)

best_regress_model <-
  results$method[which.min(results$RMSE)]

# predict values using the best model
pred_regress <-
  predict.train(
    object = regress_fits[[best_regress_model]],
    newdata = test_set_r,
    type = "raw")

pred_regress <- round(pred_regress)
```

The results from running regression analysis using various model methods is:

```{r print reg results, echo=F, results="asis"}
print_regress_results %>% knitr::kable()

```
  
```{r print accuracy, echo = F, results = "asis"}
## Accuracy

tmptext <- paste0("\n\"",
  results$method[which.min(results$RMSE)],
  "\" is the best model based on a minimum RMSE of ",
  round(min(results$RMSE), 5), ".",
  "\n\nUsing the \"",
       regress_fits[[best_regress_model]]$method,
       "\" model with test data, the accuracy is: ",
        round(accuracy(pred_regress, test_set_r$rings),
        digits = 5), ".")
cat(tmptext)
```

## Classification Analysis

```{r classification, include=F}
## split the data after adding rgrp
# using index created earlier

train_set_c <- abalone_c[-idx,]
test_set_c<- abalone_c[idx,]

if(!file.exists("class_fits.Rdata")) {
  models <- c(
    "knn",
    "lda",
    "qda",
    "naive_bayes",
    "svmLinear",
    "svmRadial",
    "gamLoess",
    "multinom",
    "rf"
  )

  control <- trainControl(method = "repeatedcv",
                          number = 10,
                          repeats = 3)
  metric <- "Accuracy"
  seed <- 3

  class_fits <- lapply(models, function(model) {
    print(model)
    set.seed(seed)
    train(
      rgrp ~ .,
      data = train_set_c,
      method = model,
      metric = metric,
      trControl = control
    )
  })
  names(class_fits) <- models
  save(class_fits, file = "class_fits.Rdata")
} else {
  load(file = "class_fits.Rdata")
}

results <- map_df(class_fits, function(model) {
  method <- model$method
  Accuracy <- model$results$Accuracy %>% min()
  return(list("method" = method, "Accuracy" = Accuracy))
})

print_class_results <- results %>%
  as_tibble() %>%
  # as_data_frame() %>%
  arrange(desc(Accuracy))

best_class <-
  results$method[which.max(results$Accuracy)]

pred_class <-
  predict.train(
    object = class_fits[[best_class]],
    newdata = test_set_c)

test_set_c$rgrp <- factor(test_set_c$rgrp)

# class_accuracy <- confusionMatrix(pred_class,
#   test_set_c$rgrp)$overall[["Accuracy"]]

```

The result of running classification analysis using various model methods is:

```{r print class results, echo=F}
print_class_results %>% knitr::kable()
```

```{r class accuracy, echo = F, results = "asis"}
## Accuracy
best_accuracy <- 
  round(accuracy(pred_class, 
                 test_set_c$rgrp),digits = 5)
tmptext <- paste0("\n\nBased on comparing accuracy",
          " for all the models, \"",
          results$method[which.max(results$Accuracy)],
          "\" is the best model with a ",
          "training data accuracy of ",
          round(max(results$Accuracy), 5), ".",
          "\n\nUsing the \"",
          class_fits[[best_class]]$method,
          "\" model with the test data,",
          " the accuracy is: ", best_accuracy)
cat(tmptext)
```


# Results
<!-- (a results section that presents the modeling results and discusses the model performance; and) -->

```{r results, echo=F, results="asis"}
tmptxt <- paste0("Regression and classification analyses were done for multiple models.  The best result was obtained with classifiction using the \"",
        class_fits[[best_class]]$method,
        "\" model by segregating rings into age groups (juvenile, adult, senior).  The resulting accuracy with test data is: ",
        # round(
        #   class_fits[[best_class]]$results$Accuracy, 4),
        best_accuracy, ".")
cat(tmptxt)
```


# Conclusion
<!-- (a conclusion section that gives a brief summary of the report, its potential impact, its limitations, and future work.) -->

Given the many variables, such as food, water characteristics, and weather patterns, which are outside the scope of this dataset, predicting rings (aka age) is difficult.  Grouping age and analyzing by classification, we could get a good approximation of the abalone rings.

Additional work could be done on tuning the models, trying other methods, or including additional data on weather, water conditions, and food availability.

\pagebreak
## Citations
The Editors of Encyclopaedia Britannica. (2020, April 23). "Abalone". Encyclopedia Britanica. `https://www.britannica.com/animal/abalone`




















